---
title: "PerfOMICS"
author: "Helena Jovic"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PerfOMICS}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup}
library("PerfOMICS")
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(knitr)
# This code has been reference from the package Testing:Introduction_TestPackage.Rmd
opts_chunk$set(fig.align = "center", 
               out.width = "90%",
               fig.width = 6, fig.height = 5.5,
               dev.args=list(pointsize=10),
               par = TRUE, # needed for setting hook 
               collapse = TRUE, # collapse input & ouput code in chunks
               warning = FALSE)
knit_hooks$set(par = function(before, options, envir)
  { if(before && options$fig.show != "none") 
       par(family = "sans", mar=c(4.1,4.1,1.1,1.1), mgp=c(3,1,0), tcl=-0.5)
})
set.seed(1) # for exact reproducibility
```

## Introduction

The purpose of `PerfOMICS` is to provide package users with athletic
performance prediction scores based on major genetic variants known to
be associated with s specific aspects of sports performance and to help
profile athletes in terms of endurance, power and speed ability,
forecast effects of athletic performance, and provide training
recommendations.

See `help(package = "PerfOMICS")` for further details and references
provided by `citation("PerfOMICS")`.

To download **PerfOMICS**, use the following commands:

``` r
require("devtools")
devtools::install_github("hjovi1/PerfOMICS", build_vignettes = TRUE)
library("PerfOMICS")
```

To list all functions available in the package:

``` r
ls("package:PerfOMICS")
```

## PerfOMICS Details

The biological data analysed by PerfOMICS are the genotypes and variants
of 23 different genes associated with athletic ability. These variants
all have researched probabilities in large populations. The data
included in this R package has been synthetically synthesized using
information on the frequencies of each genotype for the 23 genes being
analysed. A function responsible for generating a data set using the
researched genotype probabilities.

Total genotype score (TGS) reflects the additive effect of genotypes on
predicting a complex trait such as athletic performance. It is an
algorithm introduced by Williams and Folland in 2007 and is also known
as a genetic predisposition score using synthetic data of 23 genetic
markers. As of yet, there does not exist an R package that calculates
this score using Williams and Folland algorithm. PerfOMICS is a novel
tool since it provides a detailed report on an athlete's polygenic
profile as well as analyses and plots the relationships between genetic
scores, genotype scores and cardiovascular fitness from the synthesized
data. This tool can further assist in helping athletes decide their
choice of sport or athletic activity, and maximize athletic potential by
understanding implications of their genetic profile on their athletic
performance and recommended training.

## PerfOMICS Tutorial

### **Locate, store and view datasets included in package**

```{r}
library("readr")
library("stats")
library("readxl")
library("plyr")
library("graphics")
```

There are three datasets included in **`extdata`** folder in `inst` for
use in this package:

1.  \``` Ind1.xlsx` `` is a dataset containing the genetic profile of
    one individual, with two columns 'Gene' and 'Variant' and 23 rows,
    where each row corresponds to one gene.

    This dataset can be accessed at any time by running the following
    commands:

    ```{r}
    # ds1 <- system.file("extdata", "Ind1.xlsx", package = "PerfOMICS")
    # data1 <- readxl::read_excel(ds1)
    # View(data1)
    ```

2.  `` `Ind2.xlsx` `` is another dataset containing the genetic profile
    of one individual, with two columns 'Gene' and 'Variant' and 23
    rows, where each row corresponds to one gene.

    This dataset can be accessed at any time by running the following
    commands:

    ```{r warning=FALSE}
    # ds2 <- system.file("extdata", "Ind2.xlsx", package = "PerfOMICS")
    # data2 <- readxl::read_excel(ds2)
    # View(data2)
    ```

3.  `` `fake_dataset.xlsx` `` is a dataset containing genetic profiles
    of 100 individual, with 25 columns corresponding to each gene,
    'Score' and 'Fitness' and 100 rows, where each row corresponds to
    one individual. The dataset was generated using the function
    \`generateDataset\`.

    This dataset can be accessed at any time by running the following
    commands:

    ```{r}
    # ds <- system.file("extdata", "fake_dataset.xlsx", package = "PerfOMICS")
    # dataset <- readxl::read_excel(ds)
    # View(dataset)
    ```

### **Using Functions for Individual Analysis**

#### `userScore:`

This function takes in a data frame containing an individual genetic
profile of 23 gene polymorphisms and is used to calculate a genetic
score on a scale of 0 to 100 using an algorithm proposed by Williams and
Folland in 2008. The algorithm is based on researched associations
between certain variants and athletic endurance performance. A score of
0, means every gene in the individual genetic profile is unfavourable
for athletic endurance performance, whereas a score of 100 means every
gene in the genetic profile is optimal for athletic endurance
performance.

```{r warning=FALSE}
# userScore(data2)
```

#### `geneOverview:`

This function takes in a data frame containing an individual genetic
profile of 23 gene polymorphisms and is used to provide an overview of a
user's genetic profile by categorizing each gene's variant into three
buckets: unfavorable, neutral and optimal for athletic endurance
performance.

```{r warning=FALSE}
# geneOverview(data2)
```

#### `powerSpeed:`

This function takes in a data frame containing an individual genetic
profile of 23 gene polymorphisms and is used to provide users a power
speed analysis on a genetic profile by checking for matches between the
genetic profile and the optimal variants for genes associated with power
and speed ability. This function returns the associated genotypes in an
individual profile, each genotype's effect on sports performance and the
corresponding recommended training.

```{r}
# powerSpeed(data2)
```

### **Using Functions for Plotting**

#### `scatterPlot:`

This function is used to visualize a scatter plot of a variable in the
provided data set and the cardiovascular fitness level of each
individual in the data set. A regression line is added to show the trend
in the data given in the scatter plot.

```{r}
# scatterPlot(dataset)
```

#### `scoreHistogram:`

This function is used to visualize a histogram of the genetic score of
each individual in the data set and a density line.

```{r}
# scoreHistogram(dataset)
```

#### `genotypeScatter:`

This function is used to visualize the relationship between genotype
score of a particular gene in an individual and their cardiovascular
fitness level. A regression line is added to show the trend in the data
given in the scatter plot.

```{r}
# genotypeScatter(dataset,dataset$ACE)
```

### **Function for Generating a Dataset**

#### `generateDataset:`

This function is used to generate and save a dataset of entries
containing 23 gene polymorphisms associatied with athletic endurance
performance ability as well as cardiovascular fitness level. Each gene
and their variants has a researched frequency in large populations. This
dataset uses these probabilities to generate a contribed dataset
suitable for the purposes of this R package. The cardiovascular fitness
of each entry is based on the genetic score.

```{r}
# generateDataset(100)
```

### Gene Information

Run these commands to access a data frame with information on the 23
genes used for this package. Information includes: function, position,
genotypes, frequencies and sources where data was retrieved.

```{r}
# infopath <- system.file("extdata", "geneInfo.csv", package = "PerfOMICS")
# geneInfo <- readr::read_csv(infopath)
# View(geneInfo)
```

### Launch Shiny App

```{r}
# runPerfOMICS()
```

## Package Structure

| - PerfOMICS
|   \|- PerfOMICS.Rproj
|   \|- DESCRIPTION
|   \|- NAMESPACE
|   \|- LICENSE
|   \|- LICENSE.md
|   \|- README.md
|   \|- README.Rmd
|   \|- build
|     \|- vignette.rds
|   \|- data
|     \|- test.rda
|   \|- inst
|     \|- CITATION
|     \|- app.R
|     \|- extdata
|         fake_dataset.xlsx
|         geneInfo.csv
|         geneInfoReferences.csv
|         Ind1.xlsx
|         Ind2.xlsx
|         powerseed.csv
|   \|- man
|     \|- dataset.Rd
|     \|- geneOverview.Rd
|     \|- generateDataset.Rd
|     \|- genotypeScatter.Rd
|     \|- powerSpeed.Rd
|     \|- scatterPlot.Rd
|     \|- scoreHistogram.Rd
|     \|- userScore.Rd
|     \|- figures
|         README.png
|   \|- R
|     \|- data.R
|     \|- gene_overview.R
|     \|- generate_dataset.R
|     \|- powerspeed_analysis.R
|     \|- scatter_plot.R
|     \|- histo_plot.R
|     \|- genescore_analysis.R
|   \|- vignettes
|     \|- PerfOMICS.Rmd
|   \|- tests
|     \|- testthat.R
|     \|- testthat
|       \|- test-gene_overview.R
|       \|- test-generate_dataset.R
|       \|- test_powerspeed_analysis.R
|       \|- test-scatter_plot.R
|       \|- test_histo_plot.R
|       \|- test_genescore_analysis.R

## Assumptions

Genes and genomic variant selection for predicting athletic performance
depends upon the individual's interpretation of the association studies
and the strength of each association. This package should not be taken
as real medical advice. The development of the algorithms that take into
account genotypes of selected genes, assign "weight" values, threshold
levels and take into account variant to variant interactions in a
DNA-profile-to-trait relation, are proprietary and should be amendable
to patent protection, as are the gene and variant selection itself and
the individual genotype DNA profiles-to-trait interpretations.

## Package Reference

Jovic, H. (2022) PerfOMICS. Unpublished. URL
<https://github.com/hjovi1/PerfOMICSLinks> to an external site."

## Other References

[1] Kambouris, M., Ntalouka, F., Ziogas, G., & Maffulli, N. (2012).
Predictive Genomics DNA Profiling for Athletic Performance. Recent
patents on DNA & gene sequences, 6. <doi:10.2174/187221512802717321>

[2] Petr M, Thiel D, Kateřina K, Brož P, Malý T, Zahálka F, Vostatková
P, Wilk M, Chycki J, Stastny P. Speed and power-related gene
polymorphisms associated with playing position in elite soccer players.
Biol Sport. 2022 Mar;39(2):355-366. doi: 10.5114/biolsport.2022.105333.
Epub 2021 Apr 21. PMID: 35309536; PMCID: PMC8919892.

[3] Williams AG, Folland JP. Similarity of polygenic profiles limits the
potential for elite human physical performance. J Physiol. 2008 Jan
1;586(1):113-21. doi: 10.1113/jphysiol.2007.141887. Epub 2007 Sep 27.
PMID: 17901117; PMCID: PMC2375556.

[4] Ruiz, J. R., Arteta, D., Buxens, A., Artieda, M., Gómez-Gallego, F.,
Santiago, C., \... Lucia, A. (2010). Can we identify a power-oriented
polygenic profile? Journal of Applied Physiology, 108(3), 561--566.
<doi:10.1152/japplphysiol.01242.2009>

```{r}
sessionInfo()
```
