---
title: "PerfOMICS"
author: "Helena Jovic"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PerfOMICS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = "center", 
               out.width = "90%",
               fig.width = 6, fig.height = 5.5,
               dev.args=list(pointsize=10),
               par = TRUE, # needed for setting hook 
               collapse = TRUE, # collapse input & ouput code in chunks
               warning = FALSE)
knit_hooks$set(par = function(before, options, envir)
  { if(before && options$fig.show != "none") 
       par(family = "sans", mar=c(4.1,4.1,1.1,1.1), mgp=c(3,1,0), tcl=-0.5)
})
set.seed(1) # for exact reproducibility
```

## Introduction

See `help(package = "PerfOMICS")` for further details and references provided by `citation("PerfOMICS")`. To download **PerfOMICS**, use the following commands:

``` r
require("devtools")
devtools::install_github("hjovi1/PerfOMICS", build_vignettes = TRUE)
library("PerfOMICS")
```
To list all functions available in the package:
``` r
lsf.str("package:PerfOMICS")
```

## Details

The biological data analysed by PerfOMICS are the genotypes and variants of 
23 different genes associated with athletic ability. These variants all have
researched frequencies in large populations. The data in this R package has
been synthetically synthesized using information on the frequencies of each 
genotype for the 23 genes being analysed. The data set being generated can be
as large as needed.

Total genotype score (TGS) reflects the additive effect of genotypes
on predicting a complex trait such as athletic performance. It is an algorithm
introduced by Williams and Folland and is also known as a genetic predisposition
score using synthetic data of 23 genetic markers. As of yet, there does not
exist an R package that calculates this score using Williams and Folland
algorithm. PerfOMICS is a novel tool since it provides a detailed report on an 
athlete's polygenic profile as well as analyses and plots the frequencies of 
TGS from the synthesized data. This tool can further assist in helping athletes 
decide their choice of sport or athletic activity, maximize athletic potential 
through tailoring of athletic programs, provide guidelines on preventive 
measures and corrective action for athletes susceptible to sports-related 
injuries.

- Analysis (4 functions):
a. genotypeProb(): outputs a set of 23 genotypes with a respective genetic score 
b. calculateTGS(): calculates the total genotype score using the set from a.
c. athleteProfile(): generates an individualised athlete polygenetic profile.
d. generateDataSet(n): generates a dataset <1 individual polygenetic profiles
to be used for analysis and plots
e. printGeneInfo(): prints information on each gene that is being used in the
TGS calculation (name, function, variants and frequencies)

- Function for Plots (3 functions):
a. generateFreqPlot(): generates a frequency plot using a generated dataset of TGS
b. generateBoxPlot(): generates a box plot using a generated dataset of TGS

Assumptions:
Genes and genomic variant selection for predicting athletic performance depends
upon the individual\'s interpretation of the association studies and the 
strength of each association. This package should not be taken as real 
medical advice. The development of the algorithms that take into account 
genotypes of selected genes, assign \"weight\" values, threshold levels and 
take into account variant to variant interactions in a DNA-profile-to-trait 
relation, are proprietary and should be amendable to patent protection, as are 
the gene and variant selection itself and the individual genotype DNA 
profiles-to-trait interpretations.

## References
[1] Kambouris, M., Ntalouka, F., Ziogas, G., & Maffulli, N. (2012). Predictive Genomics
DNA Profiling for Athletic Performance. Recent patents on DNA & gene sequences, 6.
doi:10.2174/187221512802717321

[2] Petr M, Thiel D, Kateřina K, Brož P, Malý T, Zahálka F, Vostatková P, Wilk M, Chycki
J, Stastny P. Speed and power-related gene polymorphisms associated with playing
position in elite soccer players. Biol Sport. 2022 Mar;39(2):355-366. doi:
10.5114/biolsport.2022.105333. Epub 2021 Apr 21. PMID: 35309536; PMCID:
PMC8919892.					

[3] Williams AG, Folland JP. Similarity of polygenic profiles limits the potential for elite
human physical performance. J Physiol. 2008 Jan 1;586(1):113-21. doi:
10.1113/jphysiol.2007.141887. Epub 2007 Sep 27. PMID: 17901117; PMCID:
PMC2375556.

[4] Ruiz, J. R., Arteta, D., Buxens, A., Artieda, M., Gómez-Gallego, F., Santiago, C., \...
Lucia, A. (2010). Can we identify a power-oriented polygenic profile? Journal of Applied
Physiology, 108(3), 561--566. doi:10.1152/japplphysiol.01242.2009

```{r}
sessionInfo()
```
